---
title: "Class 11: Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The PDB database

The PDB is the main repository for biomolecular structure data.

Here we examine the contents of the PDB:

```{r}
db <- read.csv("Data Export Summary.csv", row.names=1)
head(db)
```

> Q1: Determine the percentage of structures solved by X-Ray and Electron Microscopy.  From the website what proportion of structures are protein?
 
```{r}
(db$Total/sum(db$Total)) * 100
```
 
X-ray is 89%, Electron Microscopy is 2%
 
```{r}
sum(db$Proteins)/sum(db$Total) *100
```
 
Protein makes up 93% of structures.
 
We could also try the datapasta package and copy from website, then "Addins" > "Paste as dataframe"
 
```{r}
library(datapasta)

tmp <- data.frame(stringsAsFactors=FALSE,
   Experimental.Method = c("X-Ray", "NMR", "Electron Microscopy", "Other",
                           "Multi Method", "Total"),
              Proteins = c(126880, 11062, 2277, 256, 129, 140604),
         Nucleic.Acids = c(2012, 1279, 31, 4, 5, 3331),
    Protein.NA.Complex = c(6547, 259, 800, 6, 2, 7614),
                 Other = c(8, 8, 0, 13, 1, 30),
                 Total = c(135447, 12608, 3108, 279, 137, 151579), row.names=1
)
```
 
 > Q2:  Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?
 
 There are 1157 as of 2019-05-07. See:
 http://www.rcsb.org/pdb/results/results.do?tabtoshow=Current&qrid=B9A22884
 
## Section 3: Bio3D

```{r}
library(bio3d)
```

```{r}
pdb <-read.pdb("1hsg")
pdb
```

There are 198 protein residues. The non-protein residues are Water (H2O) and the Ligand (MK1).


### Select protein residues and write new PDB file for viewing in VMD!
```{r}
#Select protein only residues
pdb_proteinonly <- atom.select(pdb, "protein", value=TRUE)
#write new pdb
write.pdb(pdb_proteinonly, file="1hsg_protein.pdb")
```

### Select ligand and write new PDB file for viewing in VMD!
```{r}
#Select drug/ligand only PDB
pdb_ligand <- atom.select(pdb, "ligand", value=TRUE)
#write new pdb
write.pdb(pdb_ligand, file="1hsg_ligand.pdb")
```

## Section 5: Adenylate Kinase structure analysis from PDB

```{r}
aa <-get.seq("1ake_A")
```

```{r}
b <-blast.pdb(aa)
```

```{r}
hits <-plot(b)
```

```{r}
ids <-c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <-get.pdb(ids, split =TRUE)
```
 
```{r}
pdbs <-pdbaln(files, fit =TRUE)
```







