---
title: "Class 8: Machine Learning"
author: "Arya Jahan"
date: "April 25, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-Means Clustering

Let's start with an example of the **kmeans()** function

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```

To Do:
- Use the kmeans() function setting k to 2 and nstart=20 
- Inspect/print the results 

```{r}
# Practice K means on example data
# Function is: kmeans(data, centers, nstart)

km <- kmeans(x,2,20)
km
```

> Q. How many points are in each cluster?
  30
  
> Q. What ‘component’ of your result object details      
  - cluster size?     
  - cluster assignment/membership?      
  - cluster center?
  
```{r}
km$size
km$cluster
km$centers
```

> Plot x colored by the kmeans cluster assignment and       
    add cluster centers as blue points


```{r}
plot(x, col=km$cluster)
points(km$centers, pch=18, col="blue", cex=3)
```

## Hierarchical Clustering Example

We must give the **hclust()** function a distance matrix, not the raw data as input

```{r}
# Distance Matrix Calculation
d <- dist(x)

# Clustering
hc <- hclust(d)
plot(hc)
abline(h=6, col="red")
```

```{r}
cutree(hc, k=2)  # Cut into k groups
```

```{r}
# Step 1. Generate some example data for clustering
x <- rbind(  
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1  
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2  
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3           
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))

colnames(x) <- c("x", "y")

# Step 2. Plot the data without clusteringplot(x)

#Step 3. Generate colors for known clusters 
#         (just so we can compare to hclust results)

col <- as.factor( rep(c("c1","c2","c3"), each=50) ) 

plot(x, col=col)
```

> Q. Use the dist(), hclust(), plot() and cutree()      
functions to return 2 and 3 clusters 

```{r}
hc2 <- hclust(dist(x))
plot(hc2)
abline(h=2, col="red")
abline(h=2.8, col="blue")
```

```{r}
# 2 Groups
gp2 <- cutree(hc2, k=2)
# 3 Groups
gp3 <- cutree(hc2, k=3)
```

```{r}
gp2
```

```{r}
gp3
```

```{r}
plot(x, col=gp2)
```

```{r}
plot(x, col=gp3)
```


```{r}
table(gp2)
```

```{r}
table(gp3)
```

```{r}
table(gp2, gp3)
```


> Q. How does this compare to your known 'col' groups?

# PCA: Principal Component Analysis

We will use the **prcomp()** function for PCA
```{r}
#First lets read our example data to work with.
## You can also download this file from the class website! 

mydata <- read.csv("https://tinyurl.com/expression-CSV",                    row.names=1) 

head(mydata, 10)
```

100 Genes in this dataset
```{r}
nrow(mydata)
```

10 Experiments for this dataset
```{r}
ncol(mydata)
```

```{r}
colnames(mydata)
```

Transpose data
```{r}
head(t(mydata))
```

Run PCA analysis on transpose of dataset
```{r}
pca <- prcomp(t(mydata), scale = TRUE)
```

PCA Plot
```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

Calculate the percent variance captured in each PC
```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

head(pca.var.per)
```


```{r}
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## A vector of colors for wt and ko samples 
colvec <- colnames(mydata) 
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,      xlab=paste0("PC1 (", pca.var.per[1], "%)"),      ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```


## Hands-On Section: UK Foods

```{r}
x <- read.csv("UK_foods.csv", row.names=1)
```

Q1: Number of Rows and Columns
```{r}
nrow(x)
ncol(x)
```

```{r}
head(x, 17)
```

Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
- I prefer the row.names argument in the read.csv function because it is simple and starts the dataset off right.

Barplots. Not Useful for differences!
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

Q3: Changing what optional argument in the above barplot() function results in the following plot?
- beside = F

### PCA to the Rescue

Transpose so that observations are rows and variables are columns
```{r}
pca <- prcomp(t(x))
summary(pca)
```

Plot PC1 v PC2
```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))

my_colors <- c("red", "dark green", "blue", "purple")
text(pca$x[,1], pca$x[,2], colnames(x), col=my_colors)
     
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```







